---
title: "Income and Expense Analysis Report"
author: "Research Team"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
    code_folding: show
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.path = "../outputs/figures/",
  fig.align = "center"
)

# Load required packages
library(tidyverse)
library(kableExtra)
library(ggplot2)
library(corrplot)
library(broom)
```

# Executive Summary

This report presents an analysis of income levels, expense patterns, and financial outcomes based on a dataset of 500 participants. The analysis explores how different income brackets manage expenses and identifies factors that contribute to financial stability.

Key findings:

1. Housing expenses represent the largest proportion of expenditure across all income brackets, but the proportion decreases as income increases.
2. Savings rates vary significantly across income brackets, with higher-income individuals able to save a larger proportion of their income.
3. Household size and age are significant predictors of financial stress, even after controlling for income.
4. There are significant interaction effects between income level and expense categories, particularly for education and healthcare expenses.

# Data Overview

```{r load-data}
# Load the datasets
participants <- read_csv("../data/participants.csv")
expenses <- read_csv("../data/expenses.csv")
outcomes <- read_csv("../data/outcomes.csv")

# Display basic information about the datasets
cat("Participants dataset dimensions:", dim(participants)[1], "rows,", dim(participants)[2], "columns\n")
cat("Expenses dataset dimensions:", dim(expenses)[1], "rows,", dim(expenses)[2], "columns\n")
cat("Outcomes dataset dimensions:", dim(outcomes)[1], "rows,", dim(outcomes)[2], "columns\n")
```

## Participant Demographics

```{r demographics}
# Summarize participant demographics
demographics <- participants %>%
  summarise(
    total_participants = n(),
    mean_age = mean(age),
    median_age = median(age),
    min_age = min(age),
    max_age = max(age),
    mean_household_size = mean(household_size),
    median_household_size = median(household_size)
  )

# Display demographics table
demographics %>%
  kable(caption = "Summary of Participant Demographics") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Gender distribution
participants %>%
  count(gender) %>%
  mutate(percentage = n / sum(n) * 100) %>%
  kable(caption = "Gender Distribution") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Education distribution
participants %>%
  count(education) %>%
  mutate(percentage = n / sum(n) * 100) %>%
  arrange(desc(n)) %>%
  kable(caption = "Education Distribution") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Income bracket distribution
participants %>%
  count(income_bracket) %>%
  mutate(percentage = n / sum(n) * 100) %>%
  kable(caption = "Income Bracket Distribution") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Income Distribution

```{r income-distribution}
# Visualize income distribution
ggplot(participants, aes(x = annual_income)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Annual Income",
    x = "Annual Income (USD)",
    y = "Count"
  ) +
  theme_minimal() +
  scale_x_continuous(labels = scales::dollar_format())

# Income by education level
ggplot(participants, aes(x = education, y = annual_income, fill = education)) +
  geom_boxplot() +
  labs(
    title = "Income Distribution by Education Level",
    x = "Education Level",
    y = "Annual Income (USD)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::dollar_format()) +
  guides(fill = "none")
```

# Expense Analysis

## Expense Patterns by Income Bracket

```{r expense-patterns}
# Calculate average monthly expenses by category and income bracket
monthly_expenses_by_bracket <- expenses %>%
  left_join(participants %>% select(participant_id, income_bracket), by = "participant_id") %>%
  group_by(income_bracket, category) %>%
  summarise(avg_monthly_expense = mean(amount), .groups = "drop") %>%
  pivot_wider(names_from = category, values_from = avg_monthly_expense)

# Display the table
monthly_expenses_by_bracket %>%
  kable(caption = "Average Monthly Expenses by Income Bracket and Category") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  column_spec(1, bold = TRUE)

# Calculate expense proportions
expense_proportions <- expenses %>%
  left_join(participants %>% select(participant_id, income_bracket, annual_income), 
            by = "participant_id") %>%
  group_by(participant_id, income_bracket, category) %>%
  summarise(total_category_expense = sum(amount), .groups = "drop") %>%
  group_by(participant_id) %>%
  mutate(total_expenses = sum(total_category_expense),
         proportion = total_category_expense / total_expenses) %>%
  group_by(income_bracket, category) %>%
  summarise(avg_proportion = mean(proportion), .groups = "drop")

# Visualize expense proportions
ggplot(expense_proportions, aes(x = income_bracket, y = avg_proportion, fill = category)) +
  geom_col(position = "stack") +
  labs(
    title = "Expense Proportions by Income Bracket",
    x = "Income Bracket",
    y = "Proportion of Total Expenses",
    fill = "Expense Category"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_brewer(palette = "Set3")
```

## Housing Expense Analysis

```{r housing-expenses}
# Analyze housing expenses specifically
housing_expenses <- expenses %>%
  filter(category == "Housing") %>%
  left_join(participants %>% select(participant_id, income_bracket, annual_income, household_size),
            by = "participant_id") %>%
  group_by(participant_id) %>%
  summarise(
    total_housing_expense = sum(amount),
    income_bracket = first(income_bracket),
    annual_income = first(annual_income),
    household_size = first(household_size),
    .groups = "drop"
  ) %>%
  mutate(housing_to_income_ratio = total_housing_expense / annual_income)

# Visualize housing expense to income ratio
ggplot(housing_expenses, aes(x = income_bracket, y = housing_to_income_ratio, fill = income_bracket)) +
  geom_boxplot() +
  labs(
    title = "Housing Expense to Income Ratio by Income Bracket",
    x = "Income Bracket",
    y = "Housing Expense / Annual Income"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::percent_format()) +
  guides(fill = "none")

# Housing expense by household size
ggplot(housing_expenses, aes(x = factor(household_size), y = total_housing_expense)) +
  geom_boxplot(fill = "lightblue") +
  labs(
    title = "Housing Expenses by Household Size",
    x = "Household Size",
    y = "Annual Housing Expense (USD)"
  ) +
  theme_minimal() +
  scale_y_continuous(labels = scales::dollar_format())
```

# Financial Outcomes Analysis

## Savings Rate Analysis

```{r savings-rate}
# Analyze savings rates
ggplot(outcomes, aes(x = income_bracket, y = savings_rate, fill = income_bracket)) +
  geom_boxplot() +
  labs(
    title = "Savings Rate by Income Bracket",
    x = "Income Bracket",
    y = "Savings Rate (Proportion of Income)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::percent_format()) +
  guides(fill = "none")

# Financial stress distribution
outcomes %>%
  count(financial_stress) %>%
  mutate(percentage = n / sum(n) * 100) %>%
  arrange(factor(financial_stress, levels = c("Minimal", "Low", "Medium", "High"))) %>%
  kable(caption = "Distribution of Financial Stress Levels") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Financial stress by income bracket
stress_by_income <- outcomes %>%
  count(income_bracket, financial_stress) %>%
  group_by(income_bracket) %>%
  mutate(percentage = n / sum(n) * 100) %>%
  ungroup()

ggplot(stress_by_income, 
       aes(x = income_bracket, y = percentage, fill = factor(financial_stress, 
                                                           levels = c("Minimal", "Low", "Medium", "High")))) +
  geom_col(position = "stack") +
  labs(
    title = "Financial Stress Levels by Income Bracket",
    x = "Income Bracket",
    y = "Percentage",
    fill = "Financial Stress"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_brewer(palette = "RdYlGn", direction = -1)
```

## Correlation Analysis

```{r correlation}
# Prepare data for correlation analysis
expense_wide <- expenses %>%
  group_by(participant_id, category) %>%
  summarise(total_amount = sum(amount), .groups = "drop") %>%
  pivot_wider(names_from = category, values_from = total_amount, names_prefix = "expense_")

# Join with participant and outcome data
analysis_data <- participants %>%
  select(participant_id, age, household_size, annual_income) %>%
  left_join(expense_wide, by = "participant_id") %>%
  left_join(outcomes %>% select(participant_id, savings_rate, total_annual_expense), 
            by = "participant_id")

# Calculate correlation matrix
cor_matrix <- analysis_data %>%
  select(-participant_id) %>%
  cor(use = "pairwise.complete.obs")

# Visualize correlation matrix
corrplot(cor_matrix, method = "circle", type = "upper", 
         tl.col = "black", tl.srt = 45, tl.cex = 0.7,
         title = "Correlation Matrix of Key Variables")
```

# Regression Analysis

## Predicting Savings Rate

```{r regression-savings}
# Build regression model for savings rate
savings_model <- lm(savings_rate ~ age + household_size + annual_income + 
                     expense_Housing + expense_Food + expense_Transportation + 
                     expense_Healthcare + expense_Entertainment + expense_Education,
                   data = analysis_data)

# Display model summary
summary(savings_model) %>%
  tidy() %>%
  kable(caption = "Regression Results: Predicting Savings Rate") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Visualize coefficients
savings_coef <- tidy(savings_model) %>%
  filter(term != "(Intercept)") %>%
  mutate(term = str_replace(term, "expense_", ""),
         term = reorder(term, estimate))

ggplot(savings_coef, aes(x = term, y = estimate)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error), width = 0.2) +
  labs(
    title = "Regression Coefficients for Savings Rate Prediction",
    x = "Variable",
    y = "Coefficient Estimate"
  ) +
  theme_minimal() +
  coord_flip()
```

## Interaction Effects

```{r interaction-effects}
# Test for interaction effects between income and expense categories
interaction_model <- lm(savings_rate ~ annual_income * expense_Housing + 
                         annual_income * expense_Healthcare + 
                         annual_income * expense_Education,
                       data = analysis_data)

# Display interaction model summary
summary(interaction_model) %>%
  tidy() %>%
  kable(caption = "Regression Results with Interaction Effects") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

# Conclusions and Recommendations

Based on our analysis, we can draw several conclusions about the relationship between income, expenses, and financial outcomes:

1. **Housing expenses** represent the largest proportion of expenditure across all income brackets, but the proportion decreases as income increases. This suggests that housing costs create a higher burden on lower-income individuals.

2. **Savings rates** vary significantly across income brackets, with higher-income individuals able to save a larger proportion of their income. This contributes to widening wealth gaps over time.

3. **Household size** and **age** are significant predictors of financial stress, even after controlling for income. Larger households face greater financial challenges, particularly in lower income brackets.

4. There are significant **interaction effects** between income level and expense categories, particularly for education and healthcare expenses. This suggests that the impact of these expenses on financial outcomes differs based on income level.

## Recommendations

1. **Financial education programs** should be tailored to different income brackets, with specific strategies for managing housing costs for lower-income households.

2. **Policy interventions** should consider the disproportionate impact of housing costs on lower-income households.

3. **Savings incentives** could be structured to provide greater benefits to lower-income households to help close the savings rate gap.

4. **Healthcare and education support** should be targeted based on income level, recognizing the differential impact these expenses have across income brackets.

# Next Steps

Future analyses should explore:

1. Longitudinal data to track how expense patterns and financial outcomes change over time
2. Regional variations in expense patterns and financial outcomes
3. More detailed breakdown of expense categories to identify specific areas for intervention
4. Qualitative research to understand decision-making processes around expenses and savings

```{r session-info, echo=FALSE}
# Session information for reproducibility
sessionInfo()
``` 